---
title: ""
bibliography: "misc/ref.bib"
csl: "misc/vancouver.csl"
output: 
  word_document:
    reference_docx: "ref.docx"
editor_options: 
  chunk_output_type: console
---

**Título**: Marcadores de regulación autonómica en respuesta a la prueba de umbral de potencia funcional en ciclistas de elite.

**Title**: Markers of autonomic regulation in response to functional power threshold testing in elite cyclists.

**Autores**: Matías Castillo-Aguilar^1,2^, Cristian Núñez-Espinosa ^2,3^\*.

::: {custom-style="Afiliation"}
^1^ Kinesiología Departamento, Universidad de Magallanes, Punta Arenas, Chile.

^2^ Chilean Austral Integrative Neurophysiology Group ([NIM-ACh](https://nimach.org)), Centro Asistencial de Docencia e Investigación (CADI-UMAG), Punta Arenas Chile.

^3^ Escuela de Medicina, Universidad de Magallanes, Punta Arenas, Chile.
:::

### \*Autor de correspondencia 

Cristian Núñez-Espinosa, School of Medicine, Magallanes University, Punta Arenas, Chile. Centro Asistencial de Docencia e Investigación CADI-UMAG, Chile. e-mail: [cristian.nunez\@umag.cl](mailto:cristian.nunez@umag.cl). Address: Avenida Bulnes 01855, Box 113-D. Phone: +56 61 2201411

\newpage

## Resumen  

**Objetivo**: Evaluar el impacto de una prueba de umbral de potencia funcional (FTP) sobre indicadores de regulación autonómica en ciclistas de alto rendimiento. **Material y métodos**: Un total de 12 ciclistas de sexo masculino (edad media 36.1 años ± 11.2) fueron reclutados. Se evaluaron parámetros de composición corporal usando bioimpedanciometría, así como variabilidad de la frecuencia cardíaca (HRV) antes y después de la aplicación de la prueba de FTP. **Resultados**: [...]. **Conclusión**: [...].

**Palabras clave**: [...].

\newpage

## Abstract 

**Objective**: To evaluate the impact of a functional power threshold test (FTP) on autonomic regulation indicators in high performance cyclists. **Material and methods**: A total of 12 male cyclists (mean age 36.1 years ± 11.2) were recruited. Body composition parameters were evaluated using bioimpedancemetry, as well as heart rate variability (HRV) before and after the application of the FTP assessment. **Results**: [...]. **Conclusion**: [...].

**Keywords**: [...].

\newpage

# Introduction

<!---- Antecedentes históricos ---->


<!---- Conceptualización del problema ---->


<!---- Hipótesis y objetivos ---->


# Material y methods

## Participants

[...]. 

## Instruments

### Instrument 1 

[...].

### Instrument 2

[...].

### Instrument 3 

[...].

### Instrument 4 

[...].


## Procedures 

[...].

[...].  

## Statistical analyses

Descriptive statistics were expressed as median and interquartile range (IQR) for continuous variables, and absolute and relative frequency (n [%]) for categorical outcomes.

In order to assess the relationship between autonomic indexes we used Spearman's rank correlation, given that the data does not follow an aproximate Gaussian distribution, assessed through graphical and analytical methods. To analyze the change in autonomic parameters in response to FTP measurements, we compute the mean difference with 95% confidence interval (CI) bias corrected and accelerated, calculated through bootstrap resampling technique, reporting as well, the bias corrected standardize mean difference (Hedges *g*) with their corresponding 95% CI.

To assess the influence of potential confounders on the autonomic response to FTP, we fitted a robust version of linear regression by iterated reweighted least squares (IRLS), assigning more weight to less extreme values and thus controlling for the influence of outliers when describing the estimated parameters of the model. To this end, the predictors were centered around their mean to interpret the intercept as the estimated response, while keeping the predictors constant and thus controlling for their influence.

All analyses were performed in the R programming language [@rlanguage], within RStudio [@rstudio]. Complementary R packages were used for analysis and plotting [@masspackage; @effectsize; @modelbased; @correlation; @ggplot2].

# Results

```{r setup, include=FALSE}
knitr::opts_chunk$set(warning = FALSE,
                      message = FALSE,
                      echo = FALSE)

library(cyclismProj) # paquete de datos
library(data.table)  # manipulación de datos
library(report)  # reporte
library(MASS) # robust lm
library(datawizard)

data("males", package = "cyclismProj") # cargamos los datos
```

## Autonomic activity and stress

When assessing the relationship between the associated variables within the athletes, we observed that a greater SNS activity and Stress index on baseline were associated with a smaller decrease in the PNS activity in response to the FTP test (baseline Stress, $\rho$ = 0.67, *p* = 0.017; baseline SNS, $\rho$ = 0.69, *p* = 0.013).

In this sense, a greater baseline PNS activity was associated with larger decreases on the PNS index in response to the FTP test ($\rho$ = -0.61, *p* = 0.037), and this decrease in ∆ PNS was associated with greater increases in the SNS activity and the stress index in response to the FTP test (∆ SNS, $\rho$ = -0.6, *p* = 0.039; ∆ Stress, $\rho$ = -0.62, *p* = 0.033). 

This is directly linked with PNS activity at post-SFT, whereas greater levels were associated with lower increases in Stress index and SNS activity levels in response to SFT test (∆ Stress, $\rho$ = -0.69, *p* = 0.014; ∆ SNS, $\rho$ = -0.77, *p* = 0.003). 

```{r, eval=FALSE}
# PNS
males[, mean_ic_boot(hrv_delta_pns, k = 2)]
males[, effectsize::hedges_g(hrv_basal_pns, hrv_post_pns, paired = TRUE)];
report_statistics(lm(hrv_delta_pns ~ 1, males))

# SNS
males[, mean_ic_boot(hrv_delta_sns, k = 2)]
males[, effectsize::hedges_g(hrv_basal_sns, hrv_post_sns, paired = TRUE)];
report_statistics(lm(hrv_delta_sns ~ 1, males))

#Stress
males[, mean_ic_boot(hrv_delta_stress)] |> print(digits = 2);
males[, effectsize::hedges_g(hrv_basal_stress, hrv_post_stress, paired = TRUE)];
report_statistics(lm(hrv_delta_stress ~ 1, males))
```

## Unadjusted autonomic response

The mean observed difference in the PNS index was -2.03 points (CI~95%~[-2.53, -1.62]), suggesting a decrease in PNS activity post-SFT test (*t*(11) = -8.34, *p* < 0.001, Hedges *g* = 2.24, CI~95%~[1.17, 3.29]), while the SNS index experienced an increase of 6.28 points (CI~95%~[4.47, 8.48], *t*(11) = 5.83, *p* < 0.001, Hedges' *g* = -1.56, CI~95%~[-2.38, -0.72]) relative to their baseline values. The Stress and SNS index tend to exhibit similar behaviour between measurements of the FTP (association between ∆SNS and ∆Stress, $\rho$ = 0.98, *p* < 0.001), so the Stress index also experienced and increase from baseline relative to the FTP measurements (mean difference = 23.1, CI~95%~[14.80, 33.27], *t*(11) = 4.67, *p* < 0.001, Hedges' *g* = -1.25, CI~95%~[-1.96, -0.50]). The utonomic variations within subjects can be seen in figure 1.

```{r fig-1, fig.cap="Figure 1. Inter-individual variations of the autonomic regulation indexes. Boxplots and errorbars with 95% CI based on bootstrap resampling around the mean and the within subjects response to the FTP test (represented by conected lines) are shown."}
knitr::include_graphics("figures/figure-1.tiff")
```

## Adjusted PNS response

```{r pns-comp, eval=FALSE}
model_simple <- hrv_delta_pns ~ center(ime)
model_full <- hrv_delta_pns ~ center(ime) + center(waist_hip_ratio)

report_parameters(x = rlm(model_simple, males, maxit = 1e5, acc = 1e-10))
report_parameters(x = rlm(model_full, males, maxit = 1e5, acc = 1e-10))
```

Rank based correlation analyses suggest that from all body composition parameters, skeletal muscle index (SMI) was the only one that was inversely associated with ∆PNS, suggesting that lower levels of SMI were related to a lower decrease in PNS activity in response to FTP test ($\rho$ = -0.69, *p* = 0.02). 

After fitting a simple linear model based on IRLS, we found that the ∆PNS changed from -2.03 points (CI~95%~[-2.53, -1.62]) in the first unadjusted comparison to -1.89 points (CI~95%~[-2.33, -1.45], *t*(9) = -9.73, *p* < 0.001) when controlling for SMI ($\beta$ = -0.71, CI~95%~[-1.53, 0.10], *t*(9) = -1.99, *p* = 0.078). 

However, and after testing the influence of other predictors in the model while still considering SMI as a predictor, we observed that the inclusion of the waist-to-hip ratio (WH~ratio~) yielded a significant effect on ∆PNS ($\beta$ = 7.90, CI~95%~[4.16, 11.63], *t*(8) = 4.88, *p* = 0.001), as well as for the effect of SMI on the latter ($\beta$ = -1.38, CI~95%~[-1.84, -0.92], *t*(8) = -6.94, *p* < 0.001). Thus, after adjusting for the effect of SMI and WH~ratio~, the estimated response of ∆PNS to the FTP test was -1.93 points (CI~95%~[-2.16, -1.70], *t*(8) = -19.15, *p* < 0.001). The final model explaining the PNS response to the FTP test is best described by the following equation:

$$
\Delta \operatorname{PNS} = -1.928 -1.379 \times \operatorname{SMI} + 7.898 \times \operatorname{{WH}_{ratio}} + \epsilon
$$

```{r sns-comp, eval=FALSE}
# IMO = (MBI = muscle-bone index)
model_simple <- rlm(formula = hrv_delta_sns ~ center(imo), 
                    data = males, maxit = 1e5, acc = 1e-10)
model_full <- rlm(formula = hrv_delta_sns ~ center(imo) * center(ftp_mean_power), 
                  data = males, maxit = 1e5, acc = 1e-10)

report_table(x = model_simple)
report_table(x = model_full)
report_parameters(x = model_simple)
report_parameters(x = model_full)
```

## Adjusted SNS response

Spearman's rank based correlation suggests a positive association between the muscle-bone index (MBI) and the ∆SNS, suggesting that greater values of MBI could be associated with greater increases in SNS activity in response to the SFT test ($\rho$ = 0.82, *p* = 0.001).

When fitting a robust linear regression with IRLS, we observed that the response of SNS activity was maintained after adjusting for MBI (Intercept = 6.20, CI~95%~[5.00, 7.40], *t*(10) = 11.53, *p* < 0.001), considering that for every 1 unit increase in MBI, we could expect an increase in 11.72 points in the SNS activity in response to the SFT test ($\beta$ = 11.72, CI~95%~[7.39, 16.04], *t*(10) = 6.04, *p* < 0.001).

Despite of previous findings in simple models, and after trying different combinations of predictors while keeping MBI in the final model, we could identify an interaction effect between the mean power achieved during the FTP test (Power^FTP^) and MBI ($\beta$ = -0.05, CI~95%~[-0.09, -4.99e-03], *t*(8) = -2.56, *p* = 0.033), considering that Power^FTP^ itself was not statistically influential on the outcome response ($\beta$ = -3.23e-03, CI~95%~[-0.02, 0.01], *t*(8) = -0.41, *p* = 0.692) while MBI was still significant, even after including Power^FTP^ in the equation ($\beta$ = 10.26, CI~95%~[8.10, 12.42], *t*(8) = 10.96, *p* < 0.001). In this sense, and after controlling for the effect of MBI and Power^FTP^, we observed that the estimated response of ∆SNS was 6.06 points (CI~95%~[5.52, 6.59], *t*(8) = 26.01, *p* < 0.001). The linear relationship between variables can be seen in figure 2. The final model that best explains the variations in ∆SNS response was the following:

$$
\Delta \operatorname{SNS} = 6.055 + 10.259 \times \operatorname{MBI} -0.003 \times \operatorname{{Power}^{FTP}} -0.050 \times (\operatorname{MBI} \times \operatorname{{Power}^{FTP}}) + \epsilon
$$

```{r fig-2, fig.cap="Figure 2. Linear response between autonomic parameters at baseline, post-SFT and the variations between these time periods. A, Interaction effect between Power^FTP^ and MBI; B, Linear relationship between HRV measurements."}
knitr::include_graphics("figures/figure-2.tiff")
```

# Discussion

[...].

[...].

[...].

# Conclusion

[...]. 

# Acknowledgment

[...].

# Conflicts of interest

[...].

# References

<div id="refs"></div>
